---
title: "r4ds_ch2_workflow_basic"
author: "Andy B. PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    theme: spacelab
---

# Summary

The following code will walk through some basics in R coding. I am not going to spend a lot of time here as I've been using R for almost 20 years. 

# Coding Basics

You can use R to do some basic math and arithmetic problems (even linear algebra)

```{r}

1 / 200 * 30

(59 + 73 + 2) / 3

sin(pi/ 2)

```

You can create new objects with the assignment operator, to complete more complicated functions

```{r}
m <- 3
x <- .04
b <- 5000

y <- m*x + b

print(y)
```

You can combine multiple elements into a vector (i.e., vectorize) with *c()*

```{r}

primes <- c(2, 3, 5, 7, 11, 13)

2 + primes
```

All R statements should be read: "object name gets value ___". 

# Comments

R will ignore anything with a '#' in front. This is useful for documenting what you are intending to do with your code. And, it helps make it reproducible.

```{r}
# create a vector of primes

primes <- c(2, 3, 5, 7, 11, 13)

# multiply your vector by the constant 69

primes * 69

```

For short pieces of one line code, you shouldn't need to comment every line. However, you should focus on *why* your code is the way it is, rather than the *how* or *what*. The reason is, the how and the what can always be deduced (if tediously); but, if you change parameters your colleagues will like to know why. 

# What's in a name?

Object names can have letters and numbers and '_' or ".' in them. You should use a single convention and stick to it. The most 'accepted' is *snake case*, which would mean all your code, variable names, object names, etc., should be 'variable_name'. Aim to keep names small and obvious. So 'x' or 'posterior' rather than, 'list_of_primes' or 'posterior_distribution' whenever possilbe. 

```{r}

this_is_a_really_long_name <- 2.5

```

# Function Calling

R has a lot of built in functions, and the freedom to create your own. 

```{r}
# this is explicit, and what you should do for reproducibility
seq(from = 1, to = 99)

# this is the same as above, where R will read in the order prescribed by the writer of the function.
# if you don't know that order, this could be confusing
seq(1, 99)
```


### Exercises

#### (1) Why does this code not work?

*Answer* Because your object is spelled wrong. 

#### (2) Just spelling errors, not going to do that. 

#### (3) What do the key commands in this question do? 

*Answer* Brings up hot key mapping. 

#### (4) Which of the plots is saved? 

*Answer* The 'my_bar_plot' because you explicit specify that it should save that object.