---
title: "r4ds_ch11_communication"
author: "Andy B. PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    theme: spacelab
---

# Summary

Your audience is not likely to share your background, or the the time and effort you've put into these data and your analyses. As such, you need to learn how to communicate clearly, confidently, concisely, efficiently, economically, and comprehensively.

## Packages you may need for this chapter

```{r}
pacman::p_load(tidyverse, scales, ggrepel, patchwork)
```

# Labels

Turning an exploratory graphic into an expository graphic is with good labels. Here, the `labs()` function will get you far. The purpose of the plot is to summarize the main or additional findings. Avoid using language that describes what the plot is (i.e., "Scatterplot"....). If you need to add more text, `subtitle()` and `caption()` are both really useful. Always useful to add more detail and description to variable names. Tell the whole story in a single plot.  

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an expection because of their light weight",
    caption = "data from fueleconomy.com"
  )
```

It's also possible to use mathematical equations as labels. If you want, look at `?plotmath()` to see all the options base r comes with. 

```{r}
df <- tibble(
  x = 1:10,
  y = cumsum(x^2)
)

ggplot(df, aes(x, y)) +
  geom_point() +
  labs(
    x = quote(x[i]),
    y = quote(sum(x[i]^2, i == 1, n))
  )

```

### Exercises 

#### 11.1a Create one plot on the fuel economy data with customized title, subtitle, caption, x, y , and color lables

```{r}
glimpse(mpg)

mpg |> 
  ggplot(aes(x = cyl, y = cty)) +
  geom_jitter(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Number of cylinders",
    y = "City fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with more cylinders",
    subtitle = "Most striking for suv and pickups",
    caption = "data from fueleconomy.com"
  )
```

#### 11.2a Re-create the following plot in the book. Note that both the colors and shapes of points vary by type of drivetrain

```{r}
mpg |> 
  ggplot(aes(x = cty, y = hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  labs(
    x = "City MPG",
    y = "Highway MPG",
    color = "Type of Drivetrain",
    shape = "Type of Drivetrain"
  )
```

# Annotations

In addition to labels for groups of variables it is also important to call out specific examples within variables (perhaps key indicators, or strategically, or just outliers). `geom_text()` is your first tool. It works like `geom_point()` except it has an additional aesthetic `label`. First, we make a new dataframe with the label information. 

```{r}
label_info <- mpg |> 
  group_by(drv) |> 
  arrange(desc(displ)) |> 
  slice_head(n = 1) |> 
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |> 
  select(displ, hwy, drv, drive_type)

label_info
```

Next, we use this new data frame to directly label the three groups to replace the legend with labels placed directly on the plot. This first plot, isn't great.  

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", vjust = "bottom"
  ) +
  theme(legend.position = "none")
```

What we really need to do is ensure that the labels don't overlap. For that, `ggrepel`

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5
  ) +
  theme(legend.position = "none")
```

You can use the same logic to highlight certain points on a plot with `geom_text_repel()`

```{r}
potential_outliers <- mpg |> 
  filter(hwy > 40 | (hwy > 20 & displ > 5))

mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text_repel(data = potential_outliers,
                  aes(label = model)) +
  geom_point(data = potential_outliers, 
             color = "firebrick2") +
  geom_point(data = potential_outliers,
             color = "firebrick2",
             size = 3, shape = "circle open")

```

You have many other tools at your fingertips:
* `geom_hline()` and `geom_vline()` to add reference lines. You can adjust their thickness (`linewidth = 2`) and white (`color = "white`) making them easy to see without taking away from the plot. 
* `geom_rect()` to draw a rectangle around a plot of interest. The boundaries are defined by xmin, xmax, ymin, and ymax. You could also use the package `ggforce` and use `geom_mark_hull` to annotate subsets of points with hulls, and I have no idea what that means. 
* `geom_segment()` with the arrow argument to draw attention to a point with an arrow: use the aesthetics x and y to define the start and stop points in conjunction with xend and yend. 
* `annotate` is another handy function. Geom's work better for a subset of data, annotate adds one or a few annotation elements to a plot. 

To demonstrate annotate, let's create some text to add to our plot. 

```{r}
trend_text <- "Larger engine sizes tend to\nhave lower fule economy." |> 
  str_wrap(width = 30)

trend_text
```

Then we add two layers of annotation: one with a label geom and the other with a segment geom. 

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "firebrick3"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "firebrick3",
    arrow = arrow(type = "closed")
  )
```

### Exercises 2

#### 11.1b Use `geom_text()` with infinite positions to place text at the four cornoers of the plot. This one sucked.

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "text", x = c(Inf, Inf, -Inf, -Inf), y = c(Inf, -Inf, Inf, -Inf),
    label = "some_text",
    hjust = c("right", "right", "left", "left"),
    vjust = c("top", "bottom", "top", "bottom")
  )
```

#### 11.2b Use `annotate()` to add a point geom in the middle of your last plot without having to create a tibble. Customize the shape, size, or color of the point.

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "text", x = c(Inf, Inf, -Inf, -Inf), y = c(Inf, -Inf, Inf, -Inf),
    label = "some_text",
    hjust = c("right", "right", "left", "left"),
    vjust = c("top", "bottom", "top", "bottom")
  ) +
  annotate(
    geom = "pointrange",
    x = 4.25, y = 28,
    xmin = 4, ymin = 30,
    xmax = 4, ymax = 30,
    color = "firebrick2",
    size = 6,
    shape = 18
  )
```
#### 11.3b How do annotation labels interact with faceting? How can you add a lael to a single facet? How can you put a different label in each facet? (Hint: think about the dataset that is being passsed to `geom_text()`).

```{r}
annotated_text <- data.frame(year = c("1999", "2008"),
                             displ = c(6, 5),
                             hwy = c(38, 40),
                             label = c("waffles", "chickenwings"))


mpg |>
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~year) +
  geom_text(data = annotated_text, label = annotated_text$label)
```

! I don't really like this answer. But, this works. Essentially, it's making an additional data.frame and specifying where you'd like the text. 

#### 11.4b What arguments to `geom_label()` control the appearance of the background box? 

```{r}
mpg |>
  ggplot(aes(x = cyl, y = hwy, label = rownames(mpg))) +
  geom_point() +
  geom_label(aes(fill = factor(cyl)), colour = "white", fontface = "bold") # fill
```

#### 11.5b What are the four arguments to `arrow()`? How do they work? Create a series of plots that demonstrate the most important functions? 

The type. 
```{r}
# angle, length, ends, type
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "firebrick3"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "firebrick3",
    arrow = arrow(type = "closed")
  )
```

The angle.
```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "firebrick3"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "firebrick3",
    arrow = arrow(angle = 20)
  )
```

The length.
```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "firebrick3"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "firebrick3",
    arrow = arrow(length = unit(1, "inches"))
  )
```

The Ends. 
```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "firebrick3"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "firebrick3",
    arrow = arrow(ends = "first")
  )
```

# Scales

The third way to make your plot communicate better is with scales. 

## Default Scales

Typically, `ggplot` will just add what it thinks are the best scales given your data. 

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))
```

Behind the scenes this is what's happening:

```{r}
# this should look the same as above
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  scale_y_continuous() +
  scale_x_continuous() +
  scale_color_discrete()
```

There are however, lot's of non-default scales. 

## Axis Ticks and Legend Keys

Collectively, these are called _guides_. Axes are for x and y guides, and legend's are for everything else. There are two keys on the legend: _breaks_ and _labels_. Breaks control the ticks, and labels control the label. 

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

You can use lables for the same reason, except you can also create breaks with `NULL` labels in the case where you can't present absolute numbers. 

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  scale_color_discrete(labels = c("4" = "4-wheel",
                                  "f" = "Front",
                                  "r" = "Rear"))
```

You can use the `scales()` package to further customize your presentations. 

```{r}
left_plot <- diamonds |> 
  ggplot(aes(x = price, y = cut)) +
  geom_boxplot(alhpa = 0.5) +
  scale_x_continuous(labels = label_dollar())

right_plot <- diamonds |> 
  ggplot(aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.5) +
  scale_x_continuous(
    labels = label_dollar(
      scale = 1/1000, suffix = "K"),
    breaks = seq(1000, 19000, by = 6000)
    )

gridExtra::grid.arrange(left_plot, right_plot, ncol = 2)

```

Another handy label function is the `label_percent()`

```{r}
diamonds |> 
  ggplot(aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percentage", labels = label_percent())
```

Another use of breaks is when you have a small dataset and you want to plot exactly where the data occur. 

```{r}
presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_x_date(name = NULL, breaks = presidential$start, date_labels = "'%y")

```

## Legend Layout

You can mostly use breaks and labels to control the axes and labels, but the theme setting will also be useful. Specifically, on where you will locate the legend on your visualization. 

```{r}

base <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

base + theme(legend.position = "right") #default

base + theme(legend.position = "left")

base + 
  theme(legend.position = "top") +
  guides(col = guide_legend(nrow = 3))

base +
  theme(legend.position = "bottom") + # how many rows you want the legend to be
  guides(col = guide_legend(nrow = 1))
```

If your plot is short and wide, then place the legend at the top of bottom. If it is tall or narrow, place on the side. You can also use `guide_leg_end()` or `guide_colorbar()` to further control your legend's display. 

```{r}
# here you specify the number of rows, and overrides the point size because we used alpha so we want to over-emphasize
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, 
                              override.aes = list(size = 4)))
```


## Replacing Scale

You can of course, just replace the complete scale. This is most likely and most often with continuous and color scales. 
```{r}
# so let's log transfrom so we can see the relationship between carat and price

left_plot <- diamonds |> 
  ggplot(aes(x = carat, y = price)) +
  geom_bin2d()

right_plot <- diamonds |> 
  ggplot(aes(x = log10(carat), y = log10(price))) +
  geom_bin2d()

gridExtra::grid.arrange(left_plot, right_plot, ncol = 2)
```

While the transformation shows the relationship more accurately, the labels are whonky, which we can fix with the scales. 

```{r}
diamonds |> 
  ggplot(aes(x = carat, y = price)) +
  geom_bin2d() +
  scale_y_log10() +
  scale_x_log10()
```

Color is also frequently customized. The default scales choose colors adjacent on the color wheel. RColorBrewer is anothe useful package to help ensure your presentations have *POP*. Also don't overlook accessibility. 

```{r}
default <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) 

fancy <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  scale_color_brewer(palette = "Set1")

gridExtra::grid.arrange(default, fancy, ncol = 2)

accessible <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_color_brewer(palette = "Set1")

gridExtra::grid.arrange(default, fancy, NULL, accessible, ncol = 2)
```


When you have a predefined mapping between values and colors, use the `scale_color_manual()`. 

```{r}
presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3"))
```

For continuous color, unlike using RColorBrewer scales where you cut a continuous variable, you may want to instead use `scale_color_gradient()` or `scale_fill_gradient()`. And if you have a diverging scale you should use or could use `scale_color_gradient2()` useful if you have a bipolar scale or a metric that goes positive and negative. You can also use the viridis package that has continuous(c), discrete(d), binned(b) palettes. 

```{r}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)

plot1 <- df |> 
  ggplot(aes(x = x, y = y)) +
  geom_hex() +
  coord_fixed() +
  labs(title = "Default, continuous", x = NULL, y = NULL)

plot2 <- df |> 
  ggplot(aes(x = x, y = y)) +
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_c() +
  labs(title = "Viridis, continuous", x = NULL, y = NULL)

plot3 <- df |> 
  ggplot(aes(x = x, y = y)) +
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_b() +
  labs(title = "Virids, binned", x = NULL, y = NULL)

gridExtra::grid.arrange(plot1, plot2, plot3, ncol = 3)

```

## Zooming

There are three ways to control the plot limits:
* Adjusting what data are plotted
* Setting the limits in each scale
* Setting xlim and ylim in coord_cartesian()

```{r}
#left

left_plot <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth()

right_plot <- mpg |>
  filter(displ >= 5 & displ <= 6 & hwy >= 10 & hwy <= 25) |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth()

gridExtra::grid.arrange(left_plot, right_plot, ncol = 2)
```
You can compare subsetting your data, vs zooming using limits and coord_cartesian.

```{r}
leftbottom_plot <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  scale_x_continuous(limits = c(5, 6)) +
  scale_y_continuous(limits = c(10, 25))

rightbottom_plot <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  coord_cartesian(xlim = c(5, 6),
                  ylim = c(10, 25))

gridExtra::grid.arrange(left_plot, right_plot, leftbottom_plot, rightbottom_plot, ncol = 2)
```

However, if you want to zoom out, or expand the range, it's better to use the limits function. So in this case below we have a different scale because we are comparing qualitatively different cars that have class differences in estimated mpg on the highway. So, tough to compare. 

```{r}
suv <- mpg |> 
  filter(class == "suv")

compact <- mpg |> 
  filter(class == "compact")

#left
suv |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() -> left_plot

#right
compact |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() -> right_plot

gridExtra::grid.arrange(left_plot, right_plot, ncol = 2)

```

So to get around this we share scales. However, this particular case you could have used faceting, that might not always be the best choice. This is essentially manual faceting. 

```{r}
x_scale <- scale_x_continuous(limits = range(mpg$displ))
y_scale <- scale_y_continuous((limits = range(mpg$hwy)))
col_scale <-  scale_color_discrete(limits = unique(mpg$drv))

#left
suv |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale -> leftbottom_plot

#right
compact |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  x_scale + 
  y_scale + 
  col_scale -> rightbottom_plot

gridExtra::grid.arrange(left_plot, right_plot, leftbottom_plot, rightbottom_plot,  ncol = 2)
```

### Exercises

#### 11.1c Why doesn't the following code override the default scale? 

_Answer_ Because we haven't mapped a color aesthetic.

```{r}
df <- tibble(
  x = rnorm(10^4),
  y = rnorm(10^4)
)

df |> 
  ggplot(aes(x = x, y = y)) +
  geom_hex() +
  scale_color_gradient(low = "white", high = "red") +
  coord_fixed()
```

#### 11.2c What is the first argument to every scale? How does it compare to `labs()`?

_Answer_ Name. You have to specify the name in labs (or at least you have more wiggly room)

#### 11.3c Change the disply of the presidential terms by:
*1 combining the two variants that customize colors and x-axis breaks
*2 improving the display of th y-axis
*3 labeling each term with the name of the president
*4 adding informative plot labels
*5 placing breaks every four years (this is trickier than it seems)

```{r}
presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_x_date(name = NULL, date_breaks = "4 years", date_labels = "'%y") +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3")) +
  scale_y_continuous(breaks = seq(34, 45, by = 1),
                     labels = presidential$name) +
  labs(
    x = "President",
    y = "Election Cycle",
    color = "Party",
    title = "Presidents terms since 1953",
    subtitle = "More Republicans have been nominated three terms in a row",
    caption = "exercise for r4ds"
  ) 
```

#### 11.4c First, create the following plot. Then, modify the code using override.aes to make the legend easier to see

```{r}
diamonds |> 
  ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 1/20)  +
  guides(color = guide_legend(override.aes = list(size = 10,
                                                  alpha = 1)))

```


# Themes

You can finally customize the nondata elements of your visualizations with themes. 

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()

```
There are eight themes included withe default 'grey' in ggplot. You can also install `ggthemes()` package and get many, many more. 

You have many options within them for creating stylistic choices. 

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  labs(
    title = "Larger engie sizeds tend to have lower fuel economy",
    caption = "Source: https://fueleconomy.gov"
  ) +
  theme(
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal",
    legend.box.background = element_rect(color = "black"),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0)
  )
```


### Exercises

#### 11.1d Pick a theme offered by the ggthemes package and apply it to the last plot you made.

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  labs(
    title = "Larger engie sizeds tend to have lower fuel economy",
    caption = "Source: https://fueleconomy.gov"
  ) +
  theme(
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal",
    legend.box.background = element_rect(color = "black"),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0)
  ) +
  ggthemes::theme_tufte()
```

#### 11.2d Make your axis labels blue and bold

```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  labs(
    title = "Larger engie sizeds tend to have lower fuel economy",
    caption = "Source: https://fueleconomy.gov"
  ) +
  theme(
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal",
    legend.box.background = element_rect(color = "black"),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0),
    axis.text = element_text(color = "dodgerblue4", face = "bold")
  ) 
```

# Layout

To this point we've only really focused on making single plot outstanding. But, often, in a report or in a presentation you will show small multiples of plots side-by-side to tell a story. Here, we use `patchwork()`. I've been using `gridExtra()` to this point, so we will switch now. 

```{r}
p1 <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  labs(title = "plot 1")

p2 <- mpg |> 
  ggplot(aes(x = drv, y = hwy)) +
  geom_boxplot() +
  labs(title = "plot 2")

p1 + p2 #patchwork adds a new functionality to the '+' operator
```

```{r}
p3 <- mpg |> 
  ggplot(aes(x = cty, y = hwy)) +
  geom_point() +
  labs(title = "plot 3")

(p1 | p3) / p2
```

`patchwork()` also allows you to collect legends from multiple plots into one common legend (which is terribly useful). Here is an example of the utility of patchwork in action. 

```{r}
p1 <- mpg |> 
  ggplot(aes(x = drv, y = cty, color = drv)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Plot 1")

p2 <- mpg |> 
  ggplot(aes(x = drv, y = hwy, color = drv)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Plot 2")

p3 <- mpg |> 
  ggplot(aes(x = cty, color = drv, fill = drv)) +
  geom_density(alpha = 0.5) +
  labs(title = "Plot 3")

p4 <- mpg |> 
  ggplot(aes(x = hwy, color = drv, fill = drv)) +
  geom_density(alpha = 0.5) +
  labs(title = "Plot 4")

p5 <- mpg |> 
  ggplot(aes(x = cty, y = hwy, color = drv)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~drv) +
  labs(title = "Plot 5")

(guide_area() / (p1 + p2) / (p3 + p4) / p5) +
  plot_annotation(
    title = "City and highway mileage for cars with different drivetrains",
    caption = "Source: https://fueleconomy.gov."
  ) +
  plot_layout(
    guides = "collect",
    heights = c(1, 3, 2, 4)
  ) &
  theme(legend.position = "top")

```

### Exercizes

#### 11.1e What happens if you omit the parentheses in the following plot layout. Can you explain why this happens? 

_Answer_ it's like order of operations. The parenthesis are specifying which plots to bundle together and the backslash is indicating which plot should be below. Without the parenthesis it thinks 2 goes above 3. 

```{r}
p1 <- mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  labs(title = "Plot 1")

p2 <- mpg |> 
  ggplot(aes(x = drv, y = hwy)) +
  geom_boxplot() +
  labs(title = "Plot 2")

p3 <- mpg |> 
  ggplot(aes(x = cty, y = hwy)) +
  geom_point() +
  labs(title = "Plot 3")

p1 | p2 / p3

```

