---
title: "r4ds_ch14_strings"
author: "Andy B. PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    theme: spacelab
---

# Summary

This is a section that I, and many others, struggle with. Learning how to extract data from strings, and how to turn data into strings. Working with character vectors. Don't think like English grammar.

# Prerequisites

```{r}
# Packages you will need
pacman::p_load(tidyverse, babynames)
```

All `stringr` functions begin with `str_` which is great coupled with R autocomplete so you can jog your memory. 

# Creating a String

```{r}
# the tidyverse grammar prefers " to ' but you can use both; this is the preferred method

string1 <- "This is a string"
string2 <- 'This is a string with a "quote" within the string'
```

If you get a '+' in your code it simply means you forgot to close something (a bracket, or in this case a quotation mark of some kind). If that happens press the `ESC` button and try again.

## Escapes

If you want to use a literal single of double quote in a string you will need to "escape" it with a `\`.

```{r}
double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'"
```

Likewise, if you'd like to use a literal `\` you will need to "escape `\`" it

```{r}
backslash <- "\\"
```

Beware! The printed representation of a string and the string are NOT the same thing.

```{r}
x <- c(single_quote, double_quote, backslash)

x

str_view(x)
```

## Raw Strings

Things can get tricky quickly if you have quotes and backslashes in the same string.

```{r}
tricky <- "double_quote <- \"\\\"\" # or '\"' single_quote <- '\\'' # \"'\""

str_view(tricky)
```

That is a lot of backslashes! So instead of toothpicking, you can use a raw string.

```{r}
tricky <- r"(double_quote <- "\"" # or '"' single_quote <- '\'' # or "'")"

str_view(tricky)
```

There are lots of options with 

* `r"[ ]"`
* `r"{ }"`
* `r"-( )"`
* `r"------( )"` 

You can add any number of slashes. 

## Other Special Characters

In addition to `\", \', and \\`, there are a handful of other special characters that may come in handy. The most common are `\n` for a new line, and `\tab` for tab. You can see the complete list of characters with `?Quotes()`

```{r}
?Quotes()

x <- c("one\ntwo", "one\ttwo", "\u00b5", "\U0001f604")

x
print(x)
writeLines(x) # this is the way it actually looks
str_view(x) # tidyverse version of above
```

### Exercises

#### 14.1a Create strings that contain the following values:
a. He said "That's amazing!"
b. \a\b\c\d
c. \\\\\\

```{r}
# a
amaze <- r"(He said "That's amazing!")"
amaze
str_view(amaze)
writeLines(amaze)

# b
abcd <- r"(\a\b\c\d)"
abcd
str_view(abcd)

# c
backslash_mania <- r"(\\\\\\)"
backslash_mania
str_view(backslash_mania)
```

#### 14.2a Create the following string your `R` sessions and print it. What happens to the specil `\u00a0"`? How does `str_view()` display it? Can you do a little Googlin' to figure out what this special character is? 

```{r}
x <- "This\u00a0is\u00a0tricky"
x
str_view(x)
writeLines(x)
```

So this is a non-breaking space character. This means you can add this between two strings and ensure there is no line break. You can achieve the same function with `strsplit()`, and `gsub()` or `paste()` functions:

```{r}
paste("Hello", "world", sep = "\u00a0")

```

# Creating Many Strings from Data

Creating strings from other strings helps you solve the problem where you have some text you wrote that you want to combine with strings from a data frame. 
In this section we will use `str_c()` with `str_glue()` and how you can use them with `mutate()`, as well as other `str_flatten()` arguments with `summarize()`


## str_c()

`str_c()` takes any number of vectors as arguments and returns a character vector
```{r}
str_c("x", "y")
str_c("x", "y", "z")
str_c("Hello ", c("John", "Susan"))
```

`str_c()` is similar to base `paste()` functions, but is designed to be used with mutate by obeying the tidyverse rules.

```{r}
df <- tibble(name = c("Flora", "David", "Terra", NA))

df |> 
  mutate(greeting = str_c("Hi ", name, "!"))
```

If you want the `NA` to behave in another way you can use the `coalesce()` function with in `str_c()` to be explicit about your NA and strings. 

```{r}
df |> 
  mutate(
    greeting1 = str_c("Hi ", coalesce(name, "you"), "!"),
    greeting2 = coalesce(str_c("Hi ", name, "!"), "Hi!")
  )
```

## str_glue()

IF you have many str_c() arguments, you may notice many "s. Insert `str_glue()`

```{r}
df |> 
  mutate(greeting = str_glue("Hi {name}!"))
```
Unfortunately, str_glue is inconsistent with the treatments of NA. If you need a regular `{` you will need to "escape" it. 

```{r}
df |> 
  mutate(greeting = str_glue("{{Hi {name}!}}"))
```

## str_flatten()

str_c and str_glue work well with mutate because their output is the same length as the input. What if you want something that returns a single string? Insert: `str_flatten()`

```{r}
str_flatten(c("x", "y", "z"))
str_flatten(c("x", "y", "z"), ", ")
str_flatten(c("x", "y", "z"), ", ", last = ", and ")
```

This makes it work well with summarize

```{r}
df <- tribble(
  ~ name, ~ fruit,
  "Carmen", "banana",
  "Carmen", "apple",
  "Marvin", "nectarine",
  "Terence", "cantaloupe",
  "Terence", "papaya",
  "Terence", "mandarin"
)

df |> 
  group_by(name) |> 
  summarize(fruits = str_flatten(fruit, ", "))
```

### Exercises

#### 14.1b Compare and contrast the results of `paste0()` with `str_c()` for the following inputs: 
```{r}
paste0("hi ", NA)
paste0(letters[1:2], letters[1:3])

str_c("hi ", NA)
#str_c(letters[1:2], letters[1:3])

str_c("hi ", coalesce("NA"))
str_c(letters[1], letters[1:3])
```

#### 14.2b What is the difference between paste() and paste0()? How can you re-create the equivalent of paste() with str_c()?

Using the collapse function. But, it doesn't work. 

```{r}
paste0(letters[1:2], letters[1:3])
paste(letters[1:2], letters[1:3])
```

#### 14.3b Convert the following experessions from str_c() to str_glue() or vice versa:


str_glue("The price of {food} is {price}")
str_c("I'm ", age, " years old and live in ", country")
str_glue("\\section{{", {title}, "}}")

# Extracting Data from Strings

Often, data is crammed together in a single string, so in this section we learn how to take data from strings with four functions:

* df |> separate_longer_delim(col, delim)
* df |> separate_longer_position(col, width)
* df |> separate_wider_delim(col, delim)
* df |> separate_wider_position(col, width)

These are built on pivot_longer and pivot_wider

## Separating into Rows

Separating strings into rows tends to be the most useful when the number of components varies from row to row. 

```{r}
df1 <- tibble(x = c("a,b,c", "d,e", "f"))

df1 |> 
  separate_longer_delim(x, delim = ",")
```

It is more rare to see the separate_longer_position but some really old data storage tables may necessitate the usage of this function.

```{r}
df2 <- tibble(x = c("1211", "131", "21"))

df2 |> 
  separate_longer_position(x, width = 1)
```
# Separating into Columns

Separating a string into columns tends to be the most useful when there are a fixed number of components in each string, and you want to spread them into columns. They have an additional step, naming the columns, so they can be trickier than their longer partners.

```{r}
df3 <- tibble(x = c("a10.1.2022", "b10.2.2011", "e15.1.2015"))

df3 |> 
  separate_wider_delim(
    x,
    delim = ".",
    names = c("code", "edition", "year")
  )
```

If a specific piece of a string is not useful you can use NA.

```{r}
df3 |> 
  separate_wider_delim(
    x,
    delim = ".",
    names = c("code", NA, "year")
  )
```

separate_wider_position worlds a bit different because you have to specify the width of the column. So, you need to give it a named integer vector, where the name gives the name, and the value gives you the width of the column. 

```{r}
df4 <- tibble(x = c("202215TX", "202122LA", "202325CA"))

df4 |> 
  separate_wider_position(
    x, 
    widths = c(year = 4, age = 2, state = 2)
  )

```

## Diagnosing Widening Problems

You need to have a fixed and known width to use the width functions. But, what if deeper in your data some have fewer or more values? For example above, what if someone is 100 or 9? `too_few()` and `too_many()` arguments are helpers.

```{r}
df <- tibble(x = c("1-1-1", "1-1-2", "1-3", "1-3-2", "1"))

# this wouldn't work normally as above, so we add too_few
df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z"),
    too_few = "debug"
  )
```

From this output you can then decide a remedy or a solution/explanation. Sometimes, it's best to just add an 'NA' for each missing piece.

```{r}
df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z"),
    too_few = "align_start"
  )
```

The same principles apply when you have too many.

```{r}
df <- tibble(x = c("1-1-1", "1-1-2", "1-3-5-6", "1-3-2", "1-3-5-7-9"))

df |> 
  separate_wider_delim(
    x, 
    delim = "-",
    names = c("x", "y", "z"),
    too_many = "debug"
  )
```

You have a few additional options with too many: silently drop or merge.

```{r}
df |> 
  separate_wider_delim(
    x, 
    delim = "-",
    names = c("x", "y", "z"),
    too_many = "drop"
  )
```

```{r}
df |> 
  separate_wider_delim(
    x, 
    delim = "-",
    names = c("x", "y", "z"),
    too_many = "merge"
  )

```

# Letters

Working with individual letters within a string. How to find the length, extract substrings, and handle long strings in plots and tables.

## Length

`str_length()` tells you the length of a string

```{r}
str_length(c("a", "R for Data Science", NA))

```

You could use this in conjunction with `count()` to find the length of baby names registered in the US. And then `filter()` to find the longest names

```{r}
babynames |> 
  count(length = str_length(name), wt = n) 

# when you look at this it appears some decisions are made about first and middle names
babynames |> 
  filter(str_length(name) == 15) |> 
  count(name, wt = n, sort = TRUE)
```

## Subsetting

You can extract parts of a string with `str_sub(string, start, end)`

```{r}
x <- c("apple", "banana", "pear")

str_sub(x, 1, 3)
str_sub(x, -3, -1)

# it won't fail or throw an error if the string is too short, it will just give you what you asked for
str_sub("a", 1, 5)
```

You could use `str_sub()` to find the first and last letters in each name.

```{r}
babynames |> 
  mutate(
    first = str_sub(name, 1, 1),
    last = str_sub(name, -1, -1)
  )
```

### Exercises

#### 14.1c When computing the distribution of the length of baby names, why did we use wt = n? 

_Answer_ If null, it counts the number of rows. If a variable, computes sum(wt) for each group. So, the wt = n will count the duplicates.

```{r}
babynames |> 
  count(length = str_length(name), wt = n) 

babynames |> 
  count(length = str_length(name)) 
```

#### 14.2c Use str_length() and str_sub() to extrat the middle letter from each baby name. What will you di if the string has an even number of characters?

```{r}
# I'm going to let the rounding rules work for themselves because I don't really care
babynames |> 
  mutate(middle = str_length(name) / 2,
         value = str_sub(name, middle, middle)
  )
```

#### 14.3c Are there any majojr trends in the length of baby names over time? What about the popularity of first and last letters?

```{r}

# on average baby names are increasing in length across time
babynames |>  
  group_by(year) |> 
  summarize(
    avg_length = mean(str_length(name))
  ) |> 
  ggplot(aes(x = year, y = avg_length)) +
  geom_col()

# first and last letters
babynames |> 
  group_by(year) |> 
  count(first = str_sub(name, 1, 1), wt = n) |> 
  top_n(1, n)

babynames |> 
  group_by(year) |> 
  count(last = str_sub(name, -1, -1), wt = n) |> 
  top_n(1, n)

```

# Non-English Text

There are only 26 letters in English and most of the people who wrote and write code do so in English. There are lots of issues (and I will never really deal with any), but here are a couple of the major concerns:

## Encoding

You need to understand how computers encode and store strings:

```{r}
charToRaw("Hadley")

charToRaw("Andrew")

```

We fortunately have UTF-8. Great for modern data, but will fail for older data. 

There are examples in the book. 

## Letter Variations

Think of accents. Some are stored as two values so "u" and then the "umlaut"

```{r}
u <- c("\u00fc", "u\u0308")

str_view(u)
str_length(u)
str_sub(u, 1, 1) #see how the first letter is different, because the accent is stored after

u[[1]] == u[[2]]

str_equal(u[[1]], u[[2]]) # this will adjust for encoding...
```

## Locale-Dependent Functions

So, here are the regional variations on a theme. "en_GB" vs "en_US" for example. R defaults to the locale your computer is set at, but if you share your code with someone in a different locale, things could get weird. 

Czech has a "ch" after the "h" in the alphabet. These are issues. I will likely never deal with them.

